Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array: inclusion of pulsar terms

### 2023-11-24 manuscript/manuscript_2/paper.tex
generate pp plot for smaller h

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. We show how to extend previous state-space analyses to include the pulsar terms. It is shown via astrophysically representative software injections in Gaussian measurement noise that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the Earth-terms-only approach.
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. A nHz GW influences the time of arrival (TOA) at the Earth of radio pulses from pulsars. A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars. In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers. Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. SMBHBs which are sufficiently close and massive may be individually resolvable, though no definitive detection has yet been made \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}. \newline

@@

\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term. In this paper we undertake two key tasks. Firstly we extend the methods presented in K24 to include the pulsar term. We show how to reparametrise the state-space model in terms of an additional pulsars-specific phase term. We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}. Secondly we quantify how the inclusion of the pulsar term affects (i) the accuracy with which the static GW parameters can be estimated, and (ii) the minimal detectable GW strain, compared to the case where the pulsar terms are omitted.  \newline

@@

There are $N$ pulsars in the array. The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$. The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, as measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation. The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW. In Section \ref{sec:psr_frequency} we outline the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:psr_measured} we outline the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:ss_params} we summarise the parameters of the model. The model is formulated in terms of pulse frequency time series as a proof-of-principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization we defer to future work.

\subsection{Spin evolution} \label{sec:psr_frequency}
The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ , 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$. Consequently, 	$f_{\rm em}^{(n)}(t)$ can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline

Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} are inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}. Such a phenomenological model is necessary as a predictive, first-principles theory of timing noise does not currently exist c.f. the multiple theorized mechanisms of timing noise (see discussion in Section 1 of \cite{KimpsonPTA}). Similar approaches, analogous to Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}. However, Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} do contain certain modelling omissions and idealisations which much be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}. These include the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and the formal interpretation of $d^2 f_{\rm p} / dt^2$ due to $\xi(t)$ in Equation \eqref{eq:frequency_evolution} not being differentiable.

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a a non-linear measurement equation:
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor, and $\Phi_0$ the phase offset of the GW with respect to some reference time. \newline

Equation \eqref{eq:g_func_trig} contains two simplifying assumptions which must be recognised:  
\begin{enumerate}
	\item The GW is a monochromatic plane wave.
	\item The pulsar locations are constant with respect to the observer 
\end{enumerate}
Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Sesana2010},
\begin{equation}
	\Delta f_{\rm gw} = 0.05 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^{8.5} M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{50 \mathrm{~nHz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic, if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \gtrsim 20$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic as a first pass in this introductory paper \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

\subsection{Static parameters of the model}\label{sec:ss_params}
For the model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured}, the set of static parameters which are specific to the pulsars in the array are
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
\end{equation}
The parameters which are specific to the GW source are 
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \psi, \delta, \alpha, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
\end{equation}
where $h_0$ is the characteristic wave strain, $\iota$ the orbital inclination, $\psi$ the polarisation angle of the wave, $\delta$ the declination of the source and $\alpha$ its right ascension. These parameters enter the model through Equation \eqref{eq:g_func_trig} where $H_{ij} = H_{ij}(h_0, \iota, \psi, \delta, \alpha)$ and $\boldsymbol{n}=\boldsymbol{n}(\delta,\alpha)$. The complete set of $7 + 5N$ static parameters is denoted $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$. While we have no prior information about the values of the members of $\boldsymbol{\theta}_{\rm gw}$, the members of $\boldsymbol{\theta}_{\rm psr}$
do have constraints from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ , \label{eq:model_evidence}
\end{equation}
and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}. For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e. $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. We refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, nested sampling and the combined workflow in the context of PTA searches for GWs. \newline

The inference model used by the Kalman filter in K24 drops the pulsar term, consistent with other PTA analyses. In this section we show how to remove this limitation and include the pulsar term in the state-space inference model passed to the Kalman filter. In Section \ref{sec:earth_term} we review how the pulsar term is dropped from typical PTA analyses and the modification to Equation \eqref{eq:g_func_trig} used in K24. In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar term for likelihood-based inference methods like nested sampling. This new parametrisation is validated with synthetic data in Section \ref{sec:rep_example1}.

>>>>

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. It is shown how to extend previous state-space analyses to include the pulsar terms. Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the previous approach.
\end{abstract}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
Nanohertz gravitational waves (GWs) are produced by the inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022}. A nHz GW influences the time of arrival (TOA) at the Earth of radio pulses from pulsars. A pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V} simultaneously measures the TOAs from multiple pulsars. In this way it is possible to detect GWs at frequencies that are prohibitively low for terrestrial interferometers. Consilient evidence from PTA observations has been presented \citep{2023ApJ...951L...8A,2023arXiv230616214A,2023ApJ...951L...6R,2023RAA....23g5024X} for a stochastic GW background, which arises from the incoherent superposition of multiple SMBHB sources \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. SMBHBs which are sufficiently close and massive are perhaps individually resolvable, though no definitive detection is yet claimed \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023,2023arXiv230616226A}. \newline

@@

\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term. In this paper we undertake two key tasks. Firstly we extend the methods presented in K24 to include the pulsar term. We show how to reparametrise the state-space model in terms of an additional pulsar-specific phase term. We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}. Secondly we quantify how the inclusion of the pulsar term affects (i) the accuracy with which the static GW parameters can be estimated, and (ii) the minimal detectable GW strain, compared to the case where the pulsar terms are omitted.  \newline

@@

There are $N$ pulsars in the array. The intrinsic spin frequency of the $n$-th pulsar, $f_{\rm p}^{(n)}(t)$, as measured in the local, freely-falling rest frame of the pulsar's centre of mass, evolves according to a stochastic differential equation. The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$. The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW. In Section \ref{sec:psr_frequency} we outline the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:psr_measured} we outline the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:ss_params} we summarise the parameters of the model. The model is formulated in terms of pulse frequency time series as a proof-of-principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, an extension we defer to future work.

\subsection{Spin evolution} \label{sec:psr_frequency}
The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ , 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ which parametrizes the noise amplitude and gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$. Consequently, 	$f_{\rm em}^{(n)}(t)$ can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline

Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} are inherently an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, mean-reverting, small-amplitude excursions around a smooth, secular trend \citep{NANOgrav2023,EPTA2023,Zic2023arXiv230616230Z}. Such a phenomenological model is necessary as a predictive, first-principles theory of timing noise does not currently exist c.f. the multiple theorized mechanisms of timing noise (see discussion in Section 1 of K24). Similar approaches, analogous to Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} have been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}. However, Equations \eqref{eq:frequency_evolution} -- \eqref{eq:spinevol} do contain certain modelling omissions and idealisations which must be recognized \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}. These include (i) the exclusion of phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G,Myers2021MNRAS.502.3113M,Meyers2021}; (ii) the exclusion of non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004}; (iii) the suitability of the white noise driver $\xi(t)$ for millisecond pulsars in PTAs; and (iv) the formal interpretation of $d^2 f_{\rm p} / dt^2$ due to $\xi(t)$ in Equation \eqref{eq:frequency_evolution} not being differentiable.

\subsection{Modulation of pulsar frequency by a GW} \label{sec:psr_measured}
In the presence of a GW, the intrinsic pulsar spin frequency for the $n$-th pulsar is related to the radio pulse frequency measured by an observer on Earth via a measurement equation:
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}\left [t-d^{(n)} \right ] g^{(n)}(t) +  \varepsilon^{(n)}(t)\ ,
	\label{eq:measurement}
\end{equation}
where $d^{(n)}$ labels the distance to the $n$-th pulsar, and $\varepsilon^{(n)}$ is a Gaussian measurement noise which satisfies 
\begin{align}
	\langle \varepsilon^{(n)}(t) \rangle &= 0 \ , \\
	\langle \varepsilon^{(n)}(t) \varepsilon^{(n')}(t') \rangle &= \sigma_{\rm m}^2 \delta_{n,n'} \delta(t - t') \ ,	\label{eq:vareps}
\end{align}
where $\sigma_{\rm m}$ is the covariance of the measurement noise at the telescope and is shared between all pulsars. The measurement function $g^{(n)}(t)$ is
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig}
\end{align}
where $[q^{(n)}]^i$ labels the $i$-th coordinate component of the $n$-th pulsar's position vector $\boldsymbol{q}^{(n)}$, $\Omega$ is the constant angular frequency of the GW, $\boldsymbol{n}$ is a unit vector specifying the direction of propagation of the GW, $H_{ij}$ is the spatial part of the gravitational plane wave amplitude tensor, and $\Phi_0$ the phase offset of the GW with respect to some reference time. \newline

\noindent Equation \eqref{eq:g_func_trig} contains two simplifying assumptions: 
\begin{enumerate}[leftmargin=2em]
	\item The GW is a monochromatic plane wave.
	\item The pulsar locations are constant with respect to the observer.
\end{enumerate}
Regarding point (i), studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves according to \citep[e.g.][]{Sesana2010},
\begin{equation}
	\Delta f_{\rm gw} = 0.05 \, \mathrm{nHz}\left(\frac{M_{\rm c}}{10^{8.5} M_{\odot}}\right)^{5 / 3}\left[\frac{f_{\rm gw}(t=t_1)}{50 \mathrm{~nHz}}\right]^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right) \ ,
	\label{eq:f_evolution}
\end{equation}
where $M_{\rm c}$ is the chirp mass of the SMBHB, $f_{\rm gw}(t=t_1)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ is the length of time over which $\Delta f_{\rm gw}$ is measured, which for PTAs is $\sim 10$ years. A source can be considered monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution $1/T_{\rm obs}$. Equation \eqref{eq:f_evolution} implies $(M_{\rm c}/10^{8.5} M_{\odot})^{5/3} [ f_{\rm gw}(t=t_1) / 50 \mathrm{~nHz}]^{11/3} (T_{\rm obs}/10 \mathrm{yr})^2 \gtrsim 20$ for $\Delta f_{\rm gw} \gtrsim 1/T_{\rm obs}$. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA composed of pulsars with a mean distance of 1.5 kpc, 78\% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array this fraction drops to 52\%; see Figure 7 in  \cite{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

@@

\subsection{Static parameters of the model}\label{sec:ss_params}
For the model described in Sections \ref{sec:psr_frequency} and \ref{sec:psr_measured}, the set of static parameters which are specific to the pulsars in the array are
\begin{equation}
	\boldsymbol{\theta}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),d^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
\end{equation}
The parameters which are specific to the GW source are 
\begin{equation}
	\boldsymbol{\theta}_{\rm gw} = \left \{h_0, \iota, \psi, \delta, \alpha, \Omega, \Phi_0 \right \} \ ,  \label{eq:params3}
\end{equation}
where $h_0$ is the characteristic wave strain, $\iota$ the orbital inclination, $\psi$ the polarisation angle of the wave, $\delta$ the declination of the source and $\alpha$ its right ascension. These parameters enter the model through Equation \eqref{eq:g_func_trig} where $H_{ij} = H_{ij}(h_0, \iota, \psi, \delta, \alpha)$ and $\boldsymbol{n}=\boldsymbol{n}(\delta,\alpha)$. The complete set of $7 + 5N$ static parameters is denoted $\boldsymbol{\theta} = \boldsymbol{\theta}_{\rm gw} \cup \boldsymbol{\theta}_{\rm psr}$. While we have no prior information about the values of the elements of $\boldsymbol{\theta}_{\rm gw}$, the elements of $\boldsymbol{\theta}_{\rm psr}$
do have constraints from electromagnetic observations; for example estimates of $d^{(n)}$ are accurate to $\sim$ 10$\%$ \citep{Cordes2002astro.ph..7156C, Verbiest2012ApJ...755...39V, Desvignes2016,Yao2017}.

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ , \label{eq:model_evidence}
\end{equation}
and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}. For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. Given a specific realisation of $\boldsymbol{Y}$, $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is solely a function of $\boldsymbol{\theta}$, i.e. $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta}) = \mathcal{L}(\boldsymbol{\theta})$. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. We refer the reader to Appendix  \ref{sec:kalman} for a review of Kalman filtering, nested sampling, and the combined workflow in the context of PTA searches for GWs. \newline

The inference model used by the Kalman filter in K24 drops the pulsar term, consistent with other PTA analyses. In this section we show how to remove this limitation and include the pulsar term in the state-space inference model passed to the Kalman filter. In Section \ref{sec:earth_term} we review how the pulsar term is dropped from typical PTA analyses and the modification to Equation \eqref{eq:g_func_trig} used in K24. In Section \ref{sec:pulsar_term2} we introduce a convenient reparametrisation of the pulsar term for likelihood-based inference methods such as nested sampling. The new parametrisation is validated with synthetic data in Section \ref{sec:rep_example1}.

\subsection{Earth term}\label{sec:earth_term}
The influence of the GW on the pulsar radio pulses is described by Equation \eqref{eq:g_func_trig}. The GW signal is composed of an ``Earth term", corresponding to the $\cos(-\Omega t + \Phi_0)$ component, and a ``pulsar term", corresponding to the $\cos \left \{-\Omega t +\Phi_0 + \Omega \left[1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right]  d^{(n)} \right \}$ component. The Earth term describes the GW phase at the observer on Earth, while the pulsar term describes the GW phase at the individual pulsar. The Earth term depends only on the GW source parameters and so is shared across all pulsars. In contrast the pulsar term is a function of $d^{(n)}$ and $\boldsymbol{q}^{(n)}$ and so varies between pulsars. \newline
